---
title: 'Mixtures Workshop: Grouped Lasso'
author: "Lizzy Gibson"
date: "July 18, 2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("grpreg")
library(grpreg)
#install.packages("janitor")
library(janitor)

#This turns off scientific notation
options(scipen = 999)
```

# Data Import and Cleaning 

First, load the dataset; clean up names as needed; and convert factors to, well, factors. 

```{r}
study_pop = read_csv("./Data/studypop.csv") %>% 
  clean_names(case = c("old_janitor")) %>% 
  mutate(bmi_cat3 = as.factor(bmi_cat3),
         edu_cat = as.factor(edu_cat),
         race_cat = as.factor(race_cat),
         male = as.factor(male)) 
```

Quick data descriptions; because of the lenght of the output, we don't execute this command here, but encourage you to!

```{r, eval = FALSE}

describe(study_pop)

```

Next we remove missing values and reorder predictors (environmental variables first, confounders second). In keeping with standard practice, we'll ln-transform the environmental exposures and the outcome. This is the dataset we'll use to illustrate variable selection methods. 

```{r}
data_lasso = study_pop %>% 
  mutate_at(vars(contains("la")), log) %>% 
  mutate(log_telomean = log(telomean)) %>% 
  dplyr::select(log_telomean, lbx074la:lbx187la, lbxd03la:lbx194la, everything(), -seqn, -telomean) %>% 
  na.omit(log_telomean) 

names(data_lasso)

dim(data_lasso)
```

Model fitting methods aren't "tidy-compliant" yet, so we'll focus on creating the relevant matrices and vectors. 

```{r}
# Create a matrix of predictors as x
x = model.matrix(log_telomean ~ ., data_lasso)[,-1]

# Extract outcome vector
y = data_lasso$log_telomean
```

Let's take a quick look at our design matrix.

```{r}
dim(x)
colnames(x)

#View(x)
```

It's helpful to note that chemical exposures appear in columns 1 to 18, and more traditional confounders appear in columns 19 to 36.

# Group definitions

Create grouping variable two ways. If there are coefficients to be included in the model without being penalized, assign them to group 0 (or "0"). This includes all covariates we want to keep in the model. Group must be a FACTOR. 

First, 3 groups = 8 non-Dioxin-like PCBs, 2 non-ortho PCBs, and TEQ (3 dioxins, 4 furans, 1 mono-ortho (Dioxin-like) PCBs). 

```{r}
group3 <- vector()
group3[grepl("lbx1|0",colnames(x))] <- "Non-Dioxin-like PCB"
group3[grepl("lbxd", colnames(x))] <- "TEQ"
group3[grepl("lbxf", colnames(x))] <- "TEQ"
group3[grepl("lbxh|p", colnames(x))] <- "Non-Ortho PCB"
group3[grepl("lbx118la", colnames(x))] <- "TEQ"
group3[grepl("pct", colnames(x))] <- "0"
group3[grepl("bcsi", colnames(x))] <- "0"
group3[grepl("bmi", colnames(x))] <- "0"
group3[grepl("edu", colnames(x))] <- "0"
group3[grepl("race", colnames(x))] <- "0"
group3[grepl("male", colnames(x))] <- "0"
group3[grepl("bxcot", colnames(x))] <- "0"
group3[grepl("age", colnames(x))] <- "0"
group3 <- as.factor(group3)

cbind(colnames(x), group3) #bind by columns
```

Second, 4 groups = 8 non-dioxin-like PCBs, 3 non-ortho PCBs, 3 dioxins, and 4 furans.

```{r}
group4 <- vector()
group4[grepl("lbx1|0",colnames(x))] <- "Non-Dioxin-like PCB"
group4[grepl("lbxd", colnames(x))] <- "Dioxin"
group4[grepl("lbxf", colnames(x))] <- "Furan"
group4[grepl("lbxh|p", colnames(x))] <- "Non-Ortho PCB"
group4[grepl("lbx118la", colnames(x))] <- "Non-Ortho PCB"
group4[grepl("pct", colnames(x))] <- "0"
group4[grepl("bcsi", colnames(x))] <- "0"
group4[grepl("bmi", colnames(x))] <- "0"
group4[grepl("edu", colnames(x))] <- "0"
group4[grepl("race", colnames(x))] <- "0"
group4[grepl("male", colnames(x))] <- "0"
group4[grepl("bxcot", colnames(x))] <- "0"
group4[grepl("age", colnames(x))] <- "0"
group4 <- as.factor(group4)

cbind(colnames(x), group4)
```

# Grouped Variable Selection

The code chunk below fits a three-group model using lasso, MCP, and SCAD penalties.

```{r models_3}
lasso_3 <- grpreg(x, y, group3, penalty = "grLasso")
MCP_3 <- grpreg(x, y, group3, penalty = "grMCP")
SCAD_3 <- grpreg(x, y, group3, penalty = "grSCAD")
```

Results are illustrated below. 

```{r}
par(mfrow = c(1,3))
plot(lasso_3)
plot(MCP_3)
plot(SCAD_3)
```

We repeat this process for the four-group categorization next.

```{r}
lasso_4 <- grpreg(x, y, group4, penalty = "grLasso")
MCP_4 <- grpreg(x, y, group4, penalty = "grMCP")
SCAD_4 <- grpreg(x, y, group4, penalty = "grSCAD")

par(mfrow = c(1,3))
plot(lasso_4)
plot(MCP_4)
plot(SCAD_4)
```

# Lasso w/ CV

As with `glmnet`, we can use CV to choose the tuning parameters. If not specified in code, ten-fold CV is the default. 

We perform this analysis first with the three-group model and the lasso penalty.

```{r cvlasso}
cv_lasso_3 <- cv.grpreg(x, y, group3, 
                        penalty = "grLasso", seed = 1988)

#Find the lambda that results in the smallest CV error
best_lasso_3 <- cv_lasso_3$lambda.min

#Fit model with cross-validated lambda
fit_lasso_3 <- grpreg(x, y, group3, penalty = "grLasso", lambda = best_lasso_3)

#Look at coefficients from model with best lambda
fit_lasso_3$beta
```

For completeness, we can repeat this with four groups. 

```{r}
#4 group model
cv_lasso_4 <- cv.grpreg(x, y, group4, penalty = "grLasso", seed = 1988)

#Find the lambda that results in the smallest CV error
best_lasso_4 <- cv_lasso_4$lambda.min

#Fit model with cross-validated lambda
fit_lasso_4 <- grpreg(x, y, group4, penalty = "grLasso", 
                      lambda = best_lasso_4)

#Look at coefficients from model with best lambda
fit_lasso_4$beta
```

## Data Visualization

```{r vizlasso}
#3 group model
lasso_beta_3 <- as_tibble(fit_lasso_3$beta, rownames = "variable") %>% 
  rename(beta = `0.0058`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4,
         method = "Grouped Lasso")

#Filter out covariates
lasso_beta_3 %>% 
  filter(group3 %in% c("TEQ", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group3, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "Lasso Coefficients for 3 Group Model", y = "Beta Coefficient", x = "Variable")

#4 group model
lasso_beta_4 <- as_tibble(fit_lasso_4$beta, rownames = "variable") %>% 
  rename(beta = `0.0059`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4)

#Filter out covariates
lasso_beta_4 %>% 
  filter(group4 %in% c("Dioxin", "Furan", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "Lasso Coefficients for 4 Group Model", y = "Beta Coefficient", x = "Variable")
```

# MCP w/ CV

We repeat the previous analyses, substituting the MCP penalty for the lasso. 

```{r cvmcp}
#3 group model

#CV to choose the best tuning parameter lambda (ten-fold default)
cv_mcp_3 <- cv.grpreg(x, y, group3,
                      penalty = "grMCP", 
                      seed = 1988)

#Find the lambda that results in the smallest CV error
best_mcp_3 <- cv_mcp_3$lambda.min

#Fit model with cross-validated lambda
fit_mcp_3 <- grpreg(x, y, group3, penalty = "grMCP", lambda = best_mcp_3)

#Look at coefficients from model with best lambda
fit_mcp_3$beta

#4 group model
#CV to choose the best tuning parameter lambda (ten-fold default)
cv_mcp_4 <- cv.grpreg(x, y, group4, penalty = "grMCP", seed = 1988)

#Find the lambda that results in the smallest CV error
best_mcp_4 <- cv_mcp_4$lambda.min

#Fit model with cross-validated lambda
fit_mcp_4 <- grpreg(x, y, group4, penalty = "grMCP", lambda = best_mcp_4)

#Look at coefficients from model with best lambda
fit_mcp_4$beta
```

## Data Visualization

```{r vizmcp}
#3 group model
mcp_beta_3 <- as_tibble(fit_mcp_3$beta, rownames = "variable") %>% 
  rename(beta = `0.0123`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4,
         method = "MCP")

#Filter out covariates
mcp_beta_3 %>% 
  filter(group3 %in% c("TEQ", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group3, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "MCP Coefficients for 3 Group Model", y = "Beta Coefficient", x = "Variable")

#4 group model
mcp_beta_4 <- as_tibble(fit_mcp_4$beta, rownames = "variable") %>% 
  rename(beta = `0.0085`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4)

#Filter out covariates
mcp_beta_4 %>% 
  filter(group4 %in% c("Dioxin", "Furan", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "MCP Coefficients for 4 Group Model", y = "Beta Coefficient", x = "Variable")
```

# SCAD w/ CV

```{r cvscad}
#3 group model

#CV to choose the best tuning parameter lambda (ten-fold default)
cv_scad_3 <- cv.grpreg(x, y, group3, 
                       penalty = "grSCAD", 
                       seed = 1988)

#Find the lambda that results in the smallest CV error
best_scad_3 <- cv_scad_3$lambda.min

#Fit model with cross-validated lambda
fit_scad_3 <- grpreg(x, y, group3, penalty = "grSCAD", lambda = best_scad_3)

#Look at coefficients from model with best lambda
fit_scad_3$beta

#4 group model
#CV to choose the best tuning parameter lambda (ten-fold default)
cv_scad_4 <- cv.grpreg(x, y, group4, penalty = "grSCAD", seed = 1988)

#Find the lambda that results in the smallest CV error
best_scad_4 <- cv_scad_4$lambda.min

#Fit model with cross-validated lambda
fit_scad_4 <- grpreg(x, y, group4, penalty = "grSCAD", lambda = best_scad_4)

#Look at coefficients from model with best lambda
fit_scad_4$beta
```

## Data Visualization

```{r vizscad}
#3 group model
scad_beta_3 <- as_tibble(fit_scad_3$beta, rownames = "variable") %>% 
  rename(beta = `0.0077`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4,
         method = "SCAD")

#Filter out covariates
scad_beta_3 %>% 
  filter(group3 %in% c("TEQ", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group3, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "SCAD Coefficients for 3 Group Model", y = "Beta Coefficient", x = "Variable")

#4 group model
scad_beta_4 <- as_tibble(fit_scad_4$beta, rownames = "variable") %>% 
  rename(beta = `0.0071`) %>% 
  filter(variable != "(Intercept)") %>% 
  mutate(group3 = group3,
         group4 = group4)

#Filter out covariates
scad_beta_4 %>% 
  filter(group4 %in% c("Dioxin", "Furan", "Non-Dioxin-like PCB", "Non-Ortho PCB")) %>% 
  ggplot(aes(x = variable, y = round(beta, 5))) + 
  geom_point() + theme_minimal() + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ group4, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + coord_flip() + 
  labs(title = "SCAD Coefficients for 4 Group Model", y = "Beta Coefficient", x = "Variable")
```

```{r create_4_plot}
grouped <- rbind(lasso_beta_3, mcp_beta_3, scad_beta_3) %>% 
  dplyr::select(-group3, -group4)

write_csv(grouped, "grouped_betas.csv")
```

## OLS from MCP Results

Lastly, we compare the results for MCP (with four groups) with the results from the corresponding OLS fit

```{r}
mcp_beta = 
  mcp_beta_4 %>% 
  filter(beta != 0) %>% 
  dplyr::select(term = variable, mcp = beta)

lm_from_mcp <- lm(log_telomean ~ lbx118la + lbxpcbla +  lbxhxcla + 
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct +
     bmi_cat3 + edu_cat + race_cat + male, data = data_lasso)

ols_beta =
  lm_from_mcp %>% 
  broom::tidy() %>% 
  dplyr::select(term, ols = estimate)

ols_v_gmcp = left_join(mcp_beta, ols_beta)
ols_v_gmcp

ggplot(ols_v_gmcp, aes(x = mcp, y = ols)) + geom_point()

```
