---
title: "Principal Component Analysis"
author: "Yanelli Nunez"
date: "8/13/2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("janitor")
library(janitor)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("ggfortify")
library(ggfortify)  
#install.packages("gridExtra")
library(gridExtra)
#install.packages("factoextra")
library(factoextra)
#install.packages("dplyr")
library(dplyr)
#install.packages("knitr")
library(knitr)
```

# Data Import and Cleaning

```{r data preping}
studypop <- read_csv("./Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>%
  mutate(bmi_cat3 = as.factor(bmi_cat3),
  edu_cat = as.factor(edu_cat),
  race_cat = as.factor(race_cat),
  male = as.factor(male))

#check dimensions of dataset
dim(studypop)

#summary statistics on dataset
summary(studypop)

#extract Non-dioxin-like PCB variables from the studypop data frame
my.x <- names(studypop)[grep("la", names(studypop))] 

#create a new data frame with my.x and convert values for the predictor variables to its natural log.
log.x <- data.frame(apply(studypop[,my.x], 2, FUN = function(x) log(x))) %>%
  setNames(paste(my.x, "l2", sep = ".")) #add suffix 12 to the predictors' name
 

#check dimensions of dataset
dim(log.x)

#summary statistics on dataset
summary(log.x)
```

# Principal Component Analysis (PCA)

## PCA using Log-Transformed Data

```{r PCA in log data}
#PCA
pca.ln  <- prcomp(log.x, scale = TRUE) #scale: a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place
summary(pca.ln)

eigenvalues_ln <- matrix(pca.ln$sdev^2) #eigenvalues
perc_variance <- round(100*matrix(pca.ln$sdev^2/sum(pca.ln$sdev^2)),1) #variance

#Summary table 
eigenvalues_ln <- cbind(1:18, eigenvalues_ln, perc_variance) 
colum_ln <- c("Principal Component", "Eigenvalues", "Percent Variance")
eigenvalues_ln <- kable(eigenvalues_ln, col.names = colum_ln)
eigenvalues_ln 


```

## Proportion of Variance Plots
The first two components explained 73% of the variance 
```{r PVE plots}

#Plots the proportion of variance explained by each component (scree plot)

pve.ln <- pca.ln$sdev^2/sum(pca.ln$sdev^2) #proportion of variance explain by each component
 

#log-transformed data
fviz_eig(pca.ln, main = "Log - Transformed data",
         xlab = "Principal component",
         ylim = c(0,70)) 


```

## Data Visualization of eigenvectors w/ Log-Transformed PCA Results
Loadings are the weights that each chemical contribute to the component. Scores are the sum of: loadings multiple by concentration of each chemical for each person. So you get a score for each person but you have a loading for each chemical

```{r}
pca.ln.ld <- as.data.frame.matrix(pca.ln$rotation)
pca.ln.ld$chem <- row.names(pca.ln.ld)

loadings_pca <- pca.ln.ld %>% 
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
plot_loadings_pca <- loadings_pca %>% 
  gather(key = "PC", value = "Loading", -chem, -Group) %>% as.tibble()

plot_loadings_pca %>% 
  filter(PC %in% c("PC1", "PC2", "PC3")) %>% 
  ggplot(aes(x = chem, y = Loading, fill = Group)) + geom_col() +
  facet_wrap(~ PC) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(title = "Loadings for First 3 Principle Components", x = "Chemicals",
       y = "Loadings")
```

## Principal Component Biplot w/ Log-Transformed PCA Results

```{r PCA biplot}
#Creates a biplot showing the loadings for principal component 1 and 2. 
autoplot(pca.ln, data = log.x, size = 0.8, colour = 'blue', alpha = 0.5,
         loadings = TRUE, loadings.colour = 'orange',
         loadings.label = TRUE, loadings.label.repel = T, 
         loadings.label.size = 2.5, loadings.label.colour = 'black',
         main = "Principal Component Analysis Biplot")
```

##Health Model

Combine PC scores with original data outcome and covariants

```{r}
x <- pca.ln$x #principal components for each chemical

pop <- read_csv("./Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>%
  dplyr::select(-(3:10), -(17:26)) %>% 
  mutate(logtelomean = log(telomean))


pop_pca <- as.tibble((cbind(pop, x)))
```

Put 2 components (continuous) into a linear regression w/ covariates to estimate association with LTL.

```{r}
health_model_pca <- lm(logtelomean ~PC1  + PC2 + PC3 +   
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_pca)


health_model_pca %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model_pca %>% broom::glance() %>% mutate(p.value = round(p.value, 5))
```

### Plot Model

```{r}
health_model_pca %>% broom::tidy() %>% as.tibble() %>% filter(grepl("PC", term)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Principal Component", title = "Association between Principal Component and LTL")

```
