---
title: "Clustering"
author: "Yanelli Nunez"
date: "8/13/2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("janitor")
library(janitor)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("ggfortify")
library(ggfortify)
#install.packages("ggdendro")
library(ggdendro)
#install.packages("ggplotify")
library(ggplotify)
#install.packages("gridExtra")
library(gridExtra) 
#install.packages("knitr")
library(knitr)
#install.packages("dendextend")
library(dendextend)
#install.packages("pryr")
library(pryr)
#install.packages("reshape2")
library(reshape2)

#This turns off scientific notation
options(scipen = 999)
```

## Data Import and Cleaning

```{r data preping}
#Need two periods to knit: "../"
#Need one period to run in console: "./"
#Or change file path to match your computer
studypop <- read_csv("../Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>%
  mutate(bmi_cat3 = as.factor(bmi_cat3),
  edu_cat = as.factor(edu_cat),
  race_cat = as.factor(race_cat),
  male = as.factor(male))

#check dimensions of dataset
dim(studypop)

#summary statistics on dataset
summary(studypop)

#extract POP variables from the studypop data frame
my.x <- names(studypop)[grep("la", names(studypop))] 

#create a new data frame with my.x and convert values for the predictor variables to its natural log.
log.x <- data.frame(apply(studypop[,my.x], 2, FUN = function(x) log(x))) %>%
  setNames(paste(my.x, "l2", sep = ".")) %>% #add suffix 12 to the predictors' name
  na.omit() #remove missing

#check dimensions of dataset
dim(log.x)

#summary statistics on dataset
summary(log.x)
```

## Heat Map

```{r}
cormat <- round(cor(studypop[grep("la", names(studypop))], use = "pairwise.complete.obs", method = c("spearman")),2)
melted_cormat <- melt(cormat) %>% rename(Correlation = value)

melted_cormat <- melted_cormat %>% 
  mutate(group3.1 = case_when(Var1 == "lbx118la" ~ "TEQ",
                           grepl("lbxf|d", Var1) == TRUE ~ "TEQ",
                           grepl("lbx1|0", Var1) == TRUE ~ "Non-Dioxin-like PCB",
                           grepl("lbxh|p", Var1) == TRUE ~ "Non-Ortho PCB")) %>% 
  mutate(group3.2 = case_when(Var2 == "lbx118la" ~ "TEQ",
                           grepl("lbxf|d", Var2) == TRUE ~ "TEQ",
                           grepl("lbx1|0", Var2) == TRUE ~ "Non-Dioxin-like PCB",
                           grepl("lbxh|p", Var2) == TRUE ~ "Non-Ortho PCB")) %>% 
  mutate(Var1 = fct_recode(Var1, "PCB 74" = "lbx074la",
                                "PCB 99" = "lbx099la",
                                "PCB 118" = "lbx118la",
                                "PCB 138" = "lbx138la",
                                "PCB 153" = "lbx153la",
                                "PCB 170" = "lbx170la",
                                "PCB 180" = "lbx180la",
                                "PCB 187" = "lbx187la",
                                "PCB 194" = "lbx194la",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la",
                               "2,3,4,7,8-pncdf" =  "lbxf03la",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la",
                               "PCB 169" =  "lbxhxcla",
                                "PCB 126" = "lbxpcbla"),
         Var2 = fct_recode(Var2, "PCB 74" = "lbx074la",
                                "PCB 99" = "lbx099la",
                                "PCB 118" = "lbx118la",
                                "PCB 138" = "lbx138la",
                                "PCB 153" = "lbx153la",
                                "PCB 170" = "lbx170la",
                                "PCB 180" = "lbx180la",
                                "PCB 187" = "lbx187la",
                                "PCB 194" = "lbx194la",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la",
                               "2,3,4,7,8-pncdf" =  "lbxf03la",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la",
                               "PCB 169" =  "lbxhxcla",
                                "PCB 126" = "lbxpcbla"))

melted_cormat %>% 
  mutate(Var1 = fct_relevel(Var1, "PCB 194", after = 8),
         Var2 = fct_relevel(Var2, "PCB 194", after = 8),
         Var1 = fct_relevel(Var1, "PCB 118", after = 8),
         Var2 = fct_relevel(Var2, "PCB 118", after = 8)) %>% 
ggplot(aes(Var1, Var2, fill = Correlation)) +
 geom_tile(color = "white") +
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                      midpoint = 0, limit = c(0,1), space = "Lab", 
                      name = "Spearman\nCorrelation") +
  theme_minimal(base_size = 15) + # minimal theme
  theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1),
        axis.text.y = element_text(vjust = 1, size = 12, hjust = 1),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15)) +
  coord_fixed() + labs(x = "", y = "")
```

```{r}
melted_cormat %>% 
 mutate(group3.1 = fct_relevel(group3.1, c("TEQ", "Non-Ortho PCB", "Non-Dioxin-like PCB"))) %>% 
 ggplot(aes(Var1, Var2, fill = Correlation)) +
 geom_tile(color = "white") +
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                      midpoint = 0, limit = c(0,1), space = "Lab", 
                      name = "Spearman\nCorrelation ") +
  scale_x_discrete(expand = c(0, 0)) + 
  scale_y_discrete(expand = c(0, 0)) + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.spacing = unit(0, "lines"),
        panel.border = element_rect(colour = "black", size=0.7),
        strip.placement = "outside", 
        legend.position = "bottom",
        strip.background = element_rect(fill = "white")) +
  labs(x = "", y = "") +
  facet_grid(group3.2 ~ group3.1, scales = "free", space = "free",
             labeller = labeller(group3.1 = label_wrap_gen(5),
                                 group3.2 = label_wrap_gen(5)))
```

## K-Means Clustering

```{r k-means}
# for reproducibility, set seed
set.seed(28)


#K-menas with K = 1
km1 <- kmeans(log.x, 1, nstart = 20)

km1$size
head(km1$centers) ## means
km1$totss ## total sum of squares
km1$withinss ## within cluster sum of squares by cluster
km1$tot.withinss ## total within cluster sum of squares -- this is what we want to minimize!
km1$betweenss ## between cluster sum of squares
Percent_bt_k1 = paste(round(100*(km1$betweenss/km1$totss),1), "%", sep = "")


#K-means with K = 2
km2 <- kmeans(log.x, 2, nstart = 20) 

km2$size
head(km2$centers)
km2$totss 
km2$withinss
km2$tot.withinss 
km2$betweenss 
Percent_bt_k2 = paste(round(100*(km2$betweenss/km2$totss),1), "%", sep = "")

#K-means with K = 3
km3 <- kmeans(log.x, 3, nstart = 20)

km3$size
head(km3$centers)
km3$totss 
km3$withinss 
km3$tot.withinss  
km3$betweenss 
Percent_bt_k3 = paste(round(100*(km3$betweenss/km3$totss),1), "%", sep = "")

#K-means with K = 5
km5 <- kmeans(log.x, 5, nstart = 20)

km5$size
head(km5$centers)
km5$totss 
km5$withinss 
km5$tot.withinss  
km5$betweenss 
Percent_bt_k5 = paste(round(100*(km3$betweenss/km3$totss),1), "%", sep = "")

#K-means with K = 10
km10 <- kmeans(log.x, 10, nstart = 20)

km10$size
head(km10$centers)
km10$totss 
km10$withinss 
km10$tot.withinss  
km10$betweenss 
Percent_bt_k10 = paste(round(100*(km10$betweenss/km10$totss),1), "%", sep = "")

#K-means with K = 100
km100 <- kmeans(log.x, 100, nstart = 20)

km100$size
head(km100$centers)
km100$totss 
km100$withinss 
km100$tot.withinss  
km100$betweenss 
Percent_bt_k100 = paste(round(100*(km100$betweenss/km100$totss),1), "%", sep = "")

```

### Total Within Cluster/Total Variation 

```{r loop, cache = TRUE}
# create an empty data frame
km.res <- data.frame(matrix(NA, (dim(log.x)[1] - 1), 3))
library(factoextra)

# use a loop to run k-means for 0 - 30 clusters and generate variance values
for (k in 1:(dim(log.x)[1] - 1)) {
  kk <- kmeans(log.x, k, nstart = 20)
  km.res[k, ] <- cbind(k, kk$tot.withinss, kk$totss) 
  }

# name the variables created
names(km.res) <- c("clusters", "WithinSS", "TotSS")

# create a variable for within cluster variance/total SS
km.res$PropWithin <- 100*km.res$WithinSS/km.res$TotSS

# compare the within cluster variance/total SS to number of clusters
km.res %>% 
  ggplot(aes(x = clusters, y = PropWithin)) + geom_line() +
  labs(title = "Optimal Number of Clusters", 
       y = "Proportion of Within over Total SS", 
       x = "Number of Clusters")

# look at a smaller subset of data
km.res[1:30,]  %>% 
  ggplot(aes(x = clusters, y = PropWithin)) + geom_point() + 
  geom_vline(xintercept = 5, color = "red", linetype = "dotted") + 
  labs(title = "Optimal Number of Clusters", 
       y = "Proportion of Within over Total SS", 
       x = "Number of Clusters")
```

### Summary Table 

```{r summary}
summary.km <- matrix(c(km1$totss, km2$totss, km3$totss, km5$totss, km10$totss, km100$totss, 
                       km1$tot.withinss, km2$tot.withinss, km3$tot.withinss, km5$tot.withinss, km10$tot.withinss, km100$tot.withinss, 
                       km1$betweenss, km2$betweenss, km3$betweenss, km5$betweenss, km10$betweenss, km100$betweenss), 
                     ncol = 3)
colnames(summary.km) <- c("Total SS", "Total within cluster SS", "Between cluster SS")
rownames(summary.km) <- c("km-1", "km-2", "km-3", "km-5", "km-10", "km-100")
summary.km <- round(as.table(summary.km), 2) 
summary.km <- knitr::kable(summary.km)
summary.km
```



k-means N = 3 using standardized data

```{r}
# for reproducibility, set seed
set.seed(29)

sd.log.x = scale(log.x)

summary(sd.log.x)

library(factoextra)
fviz_nbclust(sd.log.x, kmeans, method = "wss") ###the loop below can quickly be perform using this function. Result is the same.
fviz_nbclust(log.x, kmeans, method = "silhouette")

#K-means with K = 3
sd_km3 <- kmeans(sd.log.x, 3, nstart = 20)



```


### Visualization

```{r}

##creating data frame with the chemical means within the population
chem_means <- as.data.frame(colMeans(log.x)) #chem means
chem_means$chem <- row.names(chem_means)

plot_chem_means <- chem_means %>%
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))


##creating data fram with cluster means for k-means N=3
km3_centers <- as.data.frame.matrix(t(km3$centers)) #centers
km3_centers$chem <- row.names(km3_centers)

means_km3 <- km3_centers %>% 
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
plot_means_km3 <- means_km3 %>% 
  gather(key = "Cluster", value = "mean", -chem, -Group) %>% as.tibble()


##creating data frame with cluster means for k-means N=5
km5_centers <- as.data.frame.matrix(t(km5$centers))
km5_centers$chem <- row.names(km5_centers)

means_km5 <- km5_centers %>% 
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
plot_means_km5 <- means_km5 %>% 
  gather(key = "Cluster", value = "mean", -chem, -Group) %>% as.tibble()
```


Plots of cluster means for K-means N=3 and N=5

```{r}
 #K-means N=3
plot_chem_means %>% as.tibble() %>% rename(pop_mean = `colMeans(log.x)`) %>% 
  right_join(., plot_means_km3, by = c("Group", "chem")) %>% 
  mutate(Cluster = as.factor(Cluster),
         Cluster = fct_recode(Cluster, "Cluster 1" = "1",
         "Cluster 2" = "2",
         "Cluster 3" = "3")) %>%
  mutate(chem = fct_recode(chem, "PCB 74" = "lbx074la.l2",
                                "PCB 99" = "lbx099la.l2",
                                "PCB 118" = "lbx118la.l2",
                                "PCB 138" = "lbx138la.l2",
                                "PCB 153" = "lbx153la.l2",
                                "PCB 170" = "lbx170la.l2",
                                "PCB 180" = "lbx180la.l2",
                                "PCB 187" = "lbx187la.l2",
                                "PCB 194" = "lbx194la.l2",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la.l2",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la.l2",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la.l2",
                               "2,3,4,7,8-pncdf" =  "lbxf03la.l2",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la.l2",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la.l2",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la.l2",
                               "PCB 169" =  "lbxhxcla.l2",
                                "PCB 126" = "lbxpcbla.l2")) %>% 
  ggplot(aes(x = chem, y = mean, fill = Group)) + geom_col() +
  geom_point(aes(y = pop_mean), size = 1) +
  facet_wrap(~ Cluster) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        plot.caption = element_text(size = 11, hjust = 0),
        strip.background = element_rect(fill = "white")) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(x = "Chemicals",
       y = "Mean", caption = "*Points indicate population means")

#k-means N=5
plot_chem_means %>% as.tibble() %>% rename(pop_mean = `colMeans(log.x)`) %>% 
  right_join(., plot_means_km5, by = c("Group", "chem")) %>% 
  ggplot(aes(x = chem, y = mean, fill = Group)) + geom_col() +
  geom_point(aes(y = pop_mean), size = 1) +
  facet_wrap(~ Cluster) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        plot.caption = element_text(size = 11, hjust = 0)) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(title = "Cluster means, N = 5", x = "Chemicals",
       y = "Mean", caption = "*Points indicate population means")
```


Plots of K-means N=3 with standardized data

```{r}

##creating data fram with cluster means for k-means N=3 standarized
sd_km3_centers <- as.data.frame.matrix(t(sd_km3$centers)) #centers
sd_km3_centers$chem <- row.names(sd_km3_centers)

means_sd_km3 <- sd_km3_centers %>% 
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
plot_means_sd_km3 <- means_sd_km3 %>% 
  gather(key = "Cluster", value = "mean", -chem, -Group) %>% as.tibble()


plot_means_sd_km3 %>% 
  mutate(Cluster = as.factor(Cluster),
         Cluster = fct_recode(Cluster, "Cluster 1" = "1",
         "Cluster 2" = "2",
         "Cluster 3" = "3")) %>%
  mutate(chem = fct_recode(chem, "PCB 74" = "lbx074la.l2",
                                "PCB 99" = "lbx099la.l2",
                                "PCB 118" = "lbx118la.l2",
                                "PCB 138" = "lbx138la.l2",
                                "PCB 153" = "lbx153la.l2",
                                "PCB 170" = "lbx170la.l2",
                                "PCB 180" = "lbx180la.l2",
                                "PCB 187" = "lbx187la.l2",
                                "PCB 194" = "lbx194la.l2",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la.l2",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la.l2",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la.l2",
                               "2,3,4,7,8-pncdf" =  "lbxf03la.l2",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la.l2",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la.l2",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la.l2",
                               "PCB 169" =  "lbxhxcla.l2",
                                "PCB 126" = "lbxpcbla.l2")) %>% 
  ggplot(aes(x = chem, y = mean, fill = Group)) + geom_col() +
  facet_wrap(~ Cluster) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        plot.caption = element_text(size = 11, hjust = 0),
        strip.background = element_rect(fill = "white")) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(x = "Chemicals",
       y = "Mean",
       title = "Standardized data, K-meand, N=3")


```

### Health model 

#### 5 Cluster Model

Combine clusters grouping with original data outcome and covariants for K-means = 5

```{r}
clu.group <- km5$cluster #cluster group

#Need two periods to knit: "../"
#Need one period to run in console: "./"
#Or change file path to match your computer
pop <- read_csv("../Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>%
  select(-(3:10), -(17:26)) %>% 
  mutate(logtelomean = log(telomean))

pop_clus <- as.tibble((cbind(pop, clu.group)))
table(pop_clus$clu.group)
```

Put 5 clusters (categorical) into a linear regression w/ covariates to estimate association with LTL.

```{r}

health_model_kmeans5 <- lm(logtelomean ~ as.factor(clu.group) +
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct + age_cent + age_sq +
as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_clus)

health_model_kmeans5 %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model_kmeans5 %>% broom::glance() %>% mutate(p.value = round(p.value, 5))

confint(health_model_kmeans5)
```

##### Plot

```{r}
health_model_kmeans5 %>% broom::tidy() %>% as.tibble() %>% filter(grepl("clu.group", term)) %>% 
  mutate(term = ifelse(term == "as.factor(clu.group)2", "Cluster 2", 
                  ifelse(term == "as.factor(clu.group)3", "Cluster 3",
                    ifelse(term == "as.factor(clu.group)4", "Cluster 4", "Cluster 5")))) %>% 
      rbind(., c("Cluster 1", 0, 0, 0, 0)) %>% 
  mutate(estimate = as.numeric(estimate),
         std.error = as.numeric(std.error)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Cluster Membership", title = "Association between Clusters and LTL")

```

#### 3 Cluster Model

Combine clusters grouping with original data outcome and covariants for K-means = 3

```{r}
clu.group3 <- km3$cluster #cluster group

pop_clus3 <- as.tibble((cbind(pop, clu.group3))) %>% 
  mutate(clu.group3 = as.factor(clu.group3)) %>% 
  mutate(clu.group3 = fct_relevel(clu.group3, ref = "3"))
table(pop_clus3$clu.group3)

```

Put 3 clusters (categorical) into a linear regression w/ covariates to estimate association with LTL.

```{r}
health_model_kmeans3 <- lm(logtelomean ~ clu.group3 + age_cent + age_sq +
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_clus3)

summary(health_model_kmeans3)
confint(health_model_kmeans3)

health_model_kmeans3 %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model_kmeans3 %>% broom::glance() %>% mutate(p.value = round(p.value, 5))

#summary table of the betas and confidence intervals for the cluster variables. 

Kmeans3_betas_confint <- as.data.frame(c(health_model_kmeans3$coefficients["clu.group31"], health_model_kmeans3$coefficients["clu.group32"])) %>% 
  rownames_to_column() %>% 
  rename("variable" = 1) %>% 
  rename("beta" = 2) %>% 
  mutate('95%CI' = c("0.033, 0.132", "0.016, 0.091"),
       'P-value' = c((summary(health_model_kmeans3)$coefficients[2,4]),
                      summary(health_model_kmeans3)$coefficients[3,4]))

Kmeans3_betas_confint 
```

##### Plot

```{r}
health_model_kmeans3 %>% broom::tidy() %>% as.tibble() %>% filter(grepl("clu.group", term)) %>% 
  mutate(term = ifelse(term == "clu.group32", "Cluster 2", "Cluster 1")) %>% 
      rbind(., c("Cluster 3", 0, 0, 0, 0)) %>% 
  mutate(estimate = as.numeric(estimate),
         std.error = as.numeric(std.error)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Cluster Membership", title = "Association between Clusters and LTL")
```


## Hierarchical Clustering

```{r h clustering}
# Hierarchical clustering analysis
hc.complete <- hclust(dist(na.omit(log.x)), method = "complete")
hc.average  <- hclust(dist(na.omit(log.x)), method = "average")
hc.single   <- hclust(dist(na.omit(log.x)), method = "single")

#summary of outcome for each method
as.dendrogram(hc.complete) %>% head()
as.dendrogram(hc.average) %>% head()
as.dendrogram(hc.single) %>% head()
```

### Dendrograms

```{r dendrograms}

dend.complete <- dendro_data(hc.complete, type = "rectangle")
plot.complete <- ggplot(segment(dend.complete)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.5) +
  labs(y = "Height", x = "", title = "Complete Linkage") + theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        panel.grid = element_blank())

dend.average <- dendro_data(hc.average, type = "rectangle")
plot.average <- ggplot(segment(dend.average)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.5) +
  labs(y = "", x = "", title = "Average Linkage") + theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank())

dend.single <- dendro_data(hc.single, type = "rectangle")
plot.single <- ggplot(segment(dend.single)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.5) +
  labs(y = "", x = "", title = "Single Linkage") + theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank())

grid.arrange(plot.complete, plot.average, plot.single, ncol = 3)
```

### Cut Trees at Different Heights

```{r cutting trees}

dendro.complete <- as.dendrogram(hc.complete)

par(mfrow = c(1,1))

#cut tree at a height of 12 resulting in 2 clusters
dendro.complete %>% 
  color_branches(k = 2) %>%
  plot(main = "Complete Linkage", ylab = "Height", leaflab = "none") 
abline(h = 12, lty = 3)

#cut tree at a height of 11.25 resulting in 3 clusters
dendro.complete %>% color_branches(k = 3) %>% 
    plot(main = "Complete Linkage", ylab = "Height", leaflab = "none")
  abline(h = 11.25, lty = 3)
  
#cut tree at a height of 10 resulting in 4 clusters
dendro.complete %>% 
  color_branches(k = 4) %>%
  plot(main = "Complete Linkage", ylab = "Height", leaflab = "none") 
abline(h = 10, lty = 3)
```


### Health Models

#### 3 Cluster Model

Combine clusters grouping with original data outcome and covariants for clusters = 3

```{r}
hc.clusters.3 = cutree(hc.complete, 3)
hc_health_3 <- as.tibble(cbind(pop, hc.clusters.3)) %>% 
  mutate(hc.clusters.3 = as.factor(hc.clusters.3)) %>% 
  mutate(hc.clusters.3 = fct_relevel(hc.clusters.3, ref = "3"))

table(hc_health_3$hc.clusters.3)
```

Put 3 clusters (categorical) into a linear regression w/ covariates to estimate association with LTL.

```{r}
health_model_hc3 <- lm(logtelomean ~ hc.clusters.3 +
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct + age_cent + age_sq +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = hc_health_3)

summary(health_model_hc3)

health_model_hc3 %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model_hc3 %>% broom::glance() %>% mutate(p.value = round(p.value, 5))

confint(health_model_hc3)

#summary table of betas and confidence intervals for the cluster variables. 

hc3_betas_confint <- as.data.frame(c(health_model_hc3$coefficients["hc.clusters.31"], health_model_hc3$coefficients["hc.clusters.32"])) %>% 
  rownames_to_column() %>% 
  rename("variable" = 1) %>% 
  rename("beta" = 2) %>% 
  mutate('95%CI' = c("0.007, 0.101", "-0.002, 0.069"),
       'P-value' = c((summary(health_model_hc3)$coefficients[2,4]),
                      summary(health_model_hc3)$coefficients[3,4]))

hc3_betas_confint

```


##### Plot

Health model 

```{r}
health_model_hc3 %>% broom::tidy() %>% as.tibble() %>% filter(grepl("cluster", term)) %>% 
  mutate(term = ifelse(term == "hc.clusters.32", "Cluster 2", "Cluster 1")) %>%
      rbind(., c("Cluster 3", 0, 0, 0, 0)) %>% 
  mutate(estimate = as.numeric(estimate),
         std.error = as.numeric(std.error)) %>% 

  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Cluster Membership", title = "Association between Clusters and LTL")
```

Visualization of Hierarchical clustering N = 3 

Means of the pollutants in each cluster compare to the mean of the chemicals in the entired population to identified in which clusters individuals 
are located based on exposure. We see that in cluster 1, we find individuals that have means of chemical higher than the chemical mean in the entire population. In the case of cluster 2, most of the observations within this cluster very similar chemical concentrations to the mean of the chemical in the entire popullation. In the third cluster, we see that most of the individuals within this cluster have chemical means below the chemical means of the entire population.  

```{r}
hc3.cl <- aggregate(as.matrix(log.x) ~ as.factor(hc.clusters.3), data = hc_health_3, FUN = mean)[,-1]
 
hc3.cl <- as.data.frame.matrix(t(hc3.cl))
hc3.cl$chem <- row.names(hc3.cl)
 
means_hc3 <- hc3.cl %>%
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ",
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
 
plot_means_hc3 <- means_hc3 %>%
  rename("Cluster 3" = V1, "Cluster 2" = V2, "Cluster 1" = V3) %>% 
  gather(key = "Cluster", value = "mean", -chem, -Group) %>% as.tibble()
 
plot_chem_means %>% as.tibble() %>% rename(pop_mean = `colMeans(log.x)`) %>%
  right_join(., plot_means_hc3, by = c("Group", "chem")) %>%
  ggplot(aes(x = chem, y = mean, fill = Group)) + geom_col() +
  geom_point(aes(y = pop_mean), size = 1) +
  facet_wrap(~ Cluster) + theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        plot.caption = element_text(size = 11, hjust = 0)) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(title = "Cluster means, N = 3", x = "Chemicals",
       y = "Mean", caption = "*Points indicate population means")
```

Dendograms for hierarchical clustering using standardized data

```{r}
sd_hc.complete <- hclust(dist(na.omit(sd.log.x)), method = "complete")
sd_dend.complete <- dendro_data(sd_hc.complete, type = "rectangle")
plot.sd_complete <- ggplot(segment(sd_dend.complete)) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend), alpha = 0.5) +
  labs(y = "Height", x = "", title = "sd_Complete Linkage") + theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        panel.grid = element_blank())

sd_dendro.complete <- as.dendrogram(sd_hc.complete)
sd_dendro.complete %>% color_branches(k = 3) %>% 
    plot(main = "sd_Complete Linkage", ylab = "Height", leaflab = "none")
  abline(h = 13, lty = 3)
```

Plots for Hierarchical clustering (3 clusters) using standardized data

```{r}

sd_hc.clusters.3 = cutree(sd_hc.complete, 3)
sd_hc_health_3 <- as.tibble(cbind(pop, sd_hc.clusters.3)) %>% 
  mutate(sd_hc.clusters.3 = as.factor(sd_hc.clusters.3)) %>% 
  mutate(sd_hc.clusters.3 = fct_relevel(sd_hc.clusters.3, ref = "3"))
 
sd_hc3.cl <- aggregate(as.matrix(sd.log.x) ~ as.factor(sd_hc.clusters.3), data = sd_hc_health_3, FUN = mean)[,-1]
 
sd_hc3.cl <- as.data.frame.matrix(t(sd_hc3.cl))
sd_hc3.cl$chem <- row.names(sd_hc3.cl)
 
sd_means_hc3 <- sd_hc3.cl %>%
  mutate(Group = ifelse(chem == "lbx118la.l2", "TEQ",
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCB",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCB",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCB",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCB", "TEQ"))))))
 
plot_sd_means_hc3 <- sd_means_hc3 %>%
  rename("Cluster 3" = V1, "Cluster 2" = V2, "Cluster 1" = V3) %>% 
  gather(key = "Cluster", value = "mean", -chem, -Group) %>% as.tibble()
 
plot_sd_means_hc3 %>%
  ggplot(aes(x = chem, y = mean, fill = Group)) + geom_col() +
  facet_wrap(~ Cluster) + theme_bw() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        plot.caption = element_text(size = 11, hjust = 0)) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(title = "Cluster means, N = 3", x = "Chemicals",
       y = "Mean")
```

```{r}
health_model_sd_hc3 <- lm(logtelomean ~ hc.clusters.3 +
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct + age_cent + age_sq +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = sd_hc_health_3)

summary(health_model_sd_hc3)
```



```{r}
##compiled table for beta, CI, and P-values for the cluster coefficients of k-means and hierarchical clustering.

summary_kmeans_hc = as.data.frame(bind_rows(Kmeans3_betas_confint, hc3_betas_confint))

summary_kmeans_hc

write.csv(summary_kmeans_hc, file = "summary_kmeans_hc.csv")
```

