---
title: "Factor Analysis"
author: "Lizzy Gibson"
date: "8/9/2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
---

```{r setup2, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = "../../")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("janitor")
library(janitor)
#install.packages("psych")
library(psych)
#install.packages("GPArotation")
library(GPArotation)
#install.packages("ggrepel")
library(ggrepel)

#This turns off scientific notation
options(scipen = 999)
```

## Data Import and Cleaning

```{r clean}
studypop <- read_csv("./Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>% #remove missing
  dplyr::select(3:10, 17:26) #only include pollutants for FA, remove outcome and covariates
 

logstudypop <- log(studypop) #log transform variables
```

## Exploratory Factor Analysis (FA)

Perform an eigenvalue analysis to get an estimate (or range) for the number of underlying dimensions. Recommend fitting at least one more and one less factor than suggested by the eigenvalue analysis.

One way to determine the number of factors or components in a data or correlation matrix is to examine the “scree" plot of the successive eigenvalues. Sharp breaks in the plot suggest the appropriate number of components or factors to extract. “Parallel" analyis is an alternative technique that compares the scree of factors of the observed data with that of a random data matrix of the same size as the original.

***In factor analysis we need to pre-specify the number of factors we want. So is helpful to first run a pca to get an idea of how many factors is a good idea for the factor analysis.

```{r}
simple.pca <- princomp(logstudypop, cor = TRUE)
simple.pca$sdev^2
# 3 PC are > 1

summary(simple.pca)
# First 3 PC explain 80% of variance

```

### Orthogonal Models

Orthogonal rotation is used if it is desirable to identify factors that are as independent from one another as possible.
In Factor analysis you get values for uniqueness and comparativeness. 
```{r}
fa_2 <- fa(logstudypop, 
                 nfactors = 2, n.obs = 1003,
                 rotate = "varimax", #varimax is an orthogonal rotation
                 scores = "regression")

print(fa_2, digits = 2, sort = TRUE)

fa_3 <- fa(logstudypop, 
                 nfactors = 3, n.obs = 1003,
                 rotate = "varimax", #varimax is an orthogonal rotation
                 scores = "regression")

print(fa_3, digits = 2, sort = TRUE)

fa_4 <- fa(logstudypop, 
                 nfactors = 4, n.obs = 1003,
                 rotate = "varimax", #varimax is an orthogonal rotation
                 scores = "regression")

print(fa_4, digits = 2, sort = TRUE)

fa_5 <- fa(logstudypop, 
                 nfactors = 5, n.obs = 1003,
                 rotate = "varimax", #varimax is an orthogonal rotation
                 scores = "regression")

print(fa_5, digits = 2, sort = TRUE)
```

### Oblique Models

Oblique rotation (i.e. correlated factors) are commonly used since we often hypothesize our latent variables of interest to be correlated with one another.
Rotated ortogonal solution to get correlated results. In the orthogonal solution we only get uncorrelated results. 

```{r}
fa_2_p <- fa(logstudypop, 
                 nfactors = 2, n.obs = 1003,
                 rotate = "promax",
                 scores = "regression")

print(fa_2_p, digits = 2, sort = TRUE)

fa_3_p <- fa(logstudypop, 
                 nfactors = 3, n.obs = 1003,
                 rotate = "promax",
                 scores = "regression")

print(fa_3_p, digits = 2, sort = TRUE)

fa_4_p <- fa(logstudypop, 
                 nfactors = 4,
                 rotate = "promax",
                 scores = "regression")

print(fa_4_p, digits = 2, sort = TRUE)

fa_5_p <- fa(logstudypop, 
                 nfactors = 5, n.obs = 1003,
                 rotate = "promax",
                 scores = "regression")

print(fa_5_p, digits = 2, sort = TRUE)
```

### Fit Indices

BIC -- lower (better fit)

eBIC -- When normal theory fails (e.g., in the case of non-positive definite matrices), it useful to examine the empirically derived eBIC based upon the empirical chi^2 - 2 df.

```{r}
fit <- as.data.frame(rbind(cbind("2 Factor", "Varimax", round(fa_2$EBIC)),
      cbind("3 Factor", "Varimax", round(fa_3$EBIC)),
      cbind("4 Factor", "Varimax", round(fa_4$EBIC)),
      cbind("5 Factor", "Varimax", round(fa_5$EBIC)),
      cbind("2 Factor", "Promax", round(fa_2_p$EBIC)),
      cbind("3 Factor", "Promax", round(fa_3_p$EBIC)),
      cbind("4 Factor", "Promax", round(fa_4_p$EBIC)),
      cbind("5 Factor", "Promax", round(fa_5_p$EBIC))))

names(fit) <- c("Model", "Rotation", "EBIC")
fit %>% knitr::kable()
```

## Parameters

Choose Promax 4 factor model based on fit statistics and interpretability.

```{r}
loadings <- as.tibble(cbind(rownames(fa_4_p$loadings[]), fa_4_p$loadings[])) %>% 
  rename(Variable = V1) %>% 
  mutate(MR1 = as.numeric(MR1),
         MR2 = as.numeric(MR2),
         MR3 = as.numeric(MR3),
         MR4 = as.numeric(MR4))

loadings$Max <- colnames(loadings[,2:5])[max.col(loadings[,2:5], ties.method = "first")]

loadings %>% knitr::kable()

loadings %>% filter(Max == "MR1") %>% count()
loadings %>% filter(Max == "MR2") %>% count()
loadings %>% filter(Max == "MR3") %>% count()
loadings %>% filter(Max == "MR4") %>% count()

scores <- as.tibble(cbind(rownames(fa_4_p$scores[]), fa_4_p$scores[]))

scores$Max <- colnames(scores)[max.col(scores, ties.method = "first")]

scores

scores %>% filter(Max == "MR1") %>% count()
scores %>% filter(Max == "MR2") %>% count()
scores %>% filter(Max == "MR3") %>% count()
scores %>% filter(Max == "MR4") %>% count()

```

## Factor Correlation

```{r}
scores %>% dplyr::select(-Max) %>% corr.test()
```

## Data Visualization

```{r}
loadings <- loadings %>% 
  mutate(Group = ifelse(Variable == "lbx118la", "mPFD", 
                        ifelse(grepl("lbx1", Variable), "Non-Dioxin-like PCBs",
                          ifelse(grepl("lbx0", Variable), "Non-Dioxin-like PCBs",
                            ifelse(grepl("lbxp", Variable), "Non-Ortho PCBs",
                              ifelse(grepl("lbxh", Variable), "Non-Ortho PCBs", "mPFD"))))))

loadings %>% 
  ggplot(aes(x = MR1, y = MR2, color = Group)) + 
  geom_point() + geom_label_repel(aes(label = Variable),
                                  box.padding   = 0.35,
                                  point.padding = 0.5,
                                  segment.color = 'grey50') + 
  theme_bw() + theme(legend.position = "bottom") +
  labs(title = "Variable Loadings on First and Second Factors")

loadings %>% 
  ggplot(aes(x = MR1, y = MR3, color = Group)) + 
  geom_point() + geom_label_repel(aes(label = Variable),
                                  box.padding   = 0.35,
                                  point.padding = 0.5,
                                  segment.color = 'grey50') + 
  theme_bw() + theme(legend.position = "none") +
  labs(title = "Variable Loadings on First and Third Factors")
```

```{r}
plot_loadings <- loadings %>% 
  dplyr::select(-Max) %>% 
  gather(key = "Factor", value = "Loading", -Variable, -Group)

plot_loadings %>% 
  mutate(Factor = as.factor(Factor),
         Factor = fct_recode(Factor, "Factor 1" = "MR1",
         "Factor 2" = "MR2",
         "Factor 3" = "MR3",
         "Factor 4" = "MR4")) %>% 
  mutate(Variable = fct_recode(Variable, "PCB 74" = "lbx074la",
                                "PCB 99" = "lbx099la",
                                "PCB 118" = "lbx118la",
                                "PCB 138" = "lbx138la",
                                "PCB 153" = "lbx153la",
                                "PCB 170" = "lbx170la",
                                "PCB 180" = "lbx180la",
                                "PCB 187" = "lbx187la",
                                "PCB 194" = "lbx194la",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la",
                               "2,3,4,7,8-pncdf" =  "lbxf03la",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la",
                               "PCB 169" =  "lbxhxcla",
                                "PCB 126" = "lbxpcbla")) %>% 
  ggplot(aes(x = Variable, y = Loading, fill = Group)) + geom_col() +
  facet_wrap(~ Factor) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        strip.background = element_rect(fill = "white")) +
  coord_flip() + geom_hline(yintercept = 0, size = 0.2)

plot_loadings %>% 
  ggplot(aes(x = Factor, y = Loading, fill = Factor)) + geom_col(position = "dodge") +
  facet_wrap(~ Variable) + 
  theme_bw() + theme(legend.position = "none", axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip() + geom_hline(yintercept = 0, size = 0.2) +
  labs(title = "Variable Loadings on All Factors")
```

```{r}
scores %>% ggplot(aes(x = Max, fill = Max)) + geom_bar() +
  labs(x = "Factors", y = "Number of Individuals", title = "Number with Highest Scores per Factor") +
  theme_bw() + theme(legend.position = "none")

scores %>% group_by(Max) %>% summarise(n())

scores %>% gather(key = "factor", value = "score", MR1:MR4) %>% dplyr::select(-Max) %>% 
  ggplot(aes(x = score)) + geom_density() + facet_grid(factor~.)
```

## Health Models

Combine Factor scores with original data outcome and covariates.

```{r}
pop <- read_csv("./Data/studypop.csv") %>% 
  clean_names() %>% 
  na.omit() %>% #remove missing
  dplyr::select(-(3:10), -(17:26)) %>% 
  mutate(logtelomean = log(telomean))

pop_fa <- as.tibble(cbind(pop, scores))
```

### Continuous

Put 5 factors (continuous) into a linear regression w/ covariates to estimate association with LTL.

```{r}
health_model <- lm(logtelomean ~ MR1  + MR2 +  MR3  +  MR4 + age_cent + age_sq + 
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_fa)

summary(health_model)
confint(health_model)

health_model %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model %>% broom::glance() %>% mutate(p.value = round(p.value, 5))


#summary table of the betas and confidence intervals for the cluster variables. 

FA_betas_confint <- as.data.frame(c(health_model$coefficients["MR1"], health_model$coefficients["MR2"], health_model$coefficients["MR3"], health_model$coefficients["MR4"])) %>% rownames_to_column() %>% 
  rename("variable" = 1) %>% 
  rename("beta" = 2) %>% 
  mutate('95%CI' = c("-0.031, 0.026", "0.004, 0.050","0.006, 0.050", "-0.041, 0.000"),
       'P-value' = c((summary(health_model)$coefficients[2,4]),
                     (summary(health_model)$coefficients[3,4]),
                     (summary(health_model)$coefficients[4,4]),
                     (summary(health_model)$coefficients[4,4])))

FA_betas_confint

write.csv(FA_betas_confint, file = "FA_betas_confint.csv")
```

### Categorical

Model with 5 factors as a categorical variable (assigned to factor with highest score).

```{r}
categorical <- lm(logtelomean ~ as.factor(Max) + age_cent + age_sq + 
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_fa)

summary(categorical)
confint(categorical)

categorical %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

categorical %>% broom::glance() %>% mutate(p.value = round(p.value, 5))

```

## Plot Models

### Continuous

```{r}
health_model %>% broom::tidy() %>% as.tibble() %>% filter(grepl("MR", term)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Factor", title = "Association between Largest Factors and LTL")
```

### Categorical

```{r}
categorical %>% broom::tidy() %>% as.tibble() %>% filter(grepl("MR", term)) %>% 
  mutate(term = ifelse(term == "as.factor(Max)MR2", "MR2",
                       ifelse(term == "as.factor(Max)MR3", "MR3", "MR4"))) %>% 
  rbind(., c("MR1", 0, 0, 0, 0)) %>% 
  mutate(estimate = as.numeric(estimate),
         std.error = as.numeric(std.error)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Factor", title = "Association between Largest Factors and LTL")
```
