---
title: "Principal Component Analysis"
author: "Yanelli Nunez"
date: "8/13/2019"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("janitor")
library(janitor)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("ggcorrplot")
library(ggcorrplot)
#install.packages("ggfortify")
library(ggfortify)  
#install.packages("gridExtra")
library(gridExtra)
#install.packages("factoextra")
library(factoextra)
#install.packages("dplyr")
library(dplyr)
#install.packages("knitr")
library(knitr)

options(scipen = 999)
```

# Data Import and Cleaning

```{r data preping}
studypop <- read_csv(here::here("Data/studypop.csv")) %>% 
  clean_names() %>% 
  na.omit() %>%
  mutate(bmi_cat3 = as.factor(bmi_cat3),
  edu_cat = as.factor(edu_cat),
  race_cat = as.factor(race_cat),
  male = as.factor(male))

#check dimensions of dataset
dim(studypop)

#summary statistics on dataset
summary(studypop)

#extract POP variables from the studypop data frame
my.x <- names(studypop)[grep("la", names(studypop))] 

#create a new data frame with my.x and convert values for the predictor variables to its natural log.
log.x <- data.frame(apply(studypop[,my.x], 2, FUN = function(x) log(x))) %>%
  setNames(paste(my.x, "l2", sep = ".")) #add suffix 12 to the predictors' name

#check dimensions of dataset
dim(log.x)

#summary statistics on dataset
summary(log.x)
```

# Principal Component Analysis (PCA)

## PCA using Log-Transformed Data

```{r PCA in log data}
#PCA
pca.ln  <- prcomp(log.x, scale = TRUE) #scale: a logical value indicating whether the variables should be scaled to have unit variance before the analysis takes place
summary(pca.ln)

eigenvalues_ln <- matrix(pca.ln$sdev^2) #eigenvalues
perc_variance <- round(100*matrix(pca.ln$sdev^2/sum(pca.ln$sdev^2)),1) #variance

#Summary table 
eigenvalues_ln <- cbind(1:18, eigenvalues_ln, perc_variance) 
colum_ln <- c("Principal Component", "Eigenvalues", "Percent Variance")
eigenvalues_ln <- kable(eigenvalues_ln, col.names = colum_ln)
eigenvalues_ln 

```

## Proportion of Variance Plots
The first two components explained 73% of the variance 
```{r PVE plots}

#Plots the proportion of variance explained by each component (scree plot)

pve.ln <- pca.ln$sdev^2/sum(pca.ln$sdev^2) #proportion of variance explain by each component
 
#log-transformed data
fviz_eig(pca.ln, main = "Log - Transformed data",
         xlab = "Principal component",
         ylim = c(0,70)) 

```

## Data Visualization of eigenvectors w/ Log-Transformed PCA Results
Loadings are the weights that each chemical contribute to the component. Scores are the sum of: loadings multiply by concentration of each chemical for each person. So you get a loading for each chemical and also a total loading for each principal component (which is the sum of the chemical's loadings). You also get a score for each person (each observation) which is the sum of the scores of each chemical (loading*chemical concentration). So for each person (observation) you have a score for each principal component. Each principal component also has a score which is the sum of the scores within the principal component. 

```{r}
pca.ln.ld <- as.data.frame.matrix(pca.ln$rotation) ## rotation is the loadings variable within the pca output.
pca.ln.ld$chem <- row.names(pca.ln.ld)

loadings_pca <- pca.ln.ld %>% 
  mutate(Group = ifelse(chem == "lbx118la.l2", "mPFD", 
                        ifelse(grepl("lbx1", chem), "Non-Dioxin-like PCBs",
                               ifelse(grepl("lbx0", chem), "Non-Dioxin-like PCBs",
                                      ifelse(grepl("lbxp", chem), "Non-Ortho PCBs",
                                             ifelse(grepl("lbxh", chem), "Non-Ortho PCBs", "mPFD"))))))
plot_loadings_pca <- loadings_pca %>% 
  gather(key = "PC", value = "Loading", -chem, -Group) %>% as.tibble()

plot_loadings_pca %>% 
  filter(PC %in% c("PC1", "PC2", "PC3")) %>% 
  mutate(PC = as.factor(PC),
         PC = fct_recode(PC, "PC 1" = "PC1",
         "PC 2" = "PC2",
         "PC 3" = "PC3")) %>% 
   mutate(chem = fct_recode(chem, "PCB 74" = "lbx074la.l2",
                                "PCB 99" = "lbx099la.l2",
                                "PCB 118" = "lbx118la.l2",
                                "PCB 138" = "lbx138la.l2",
                                "PCB 153" = "lbx153la.l2",
                                "PCB 170" = "lbx170la.l2",
                                "PCB 180" = "lbx180la.l2",
                                "PCB 187" = "lbx187la.l2",
                                "PCB 194" = "lbx194la.l2",
                                "1,2,3,6,7,8-hxcdd" = "lbxd03la.l2",
                                "1,2,3,4,6,7,8-hpcdd" = "lbxd05la.l2",
                               "1,2,3,4,6,7,8,9-ocdd" =  "lbxd07la.l2",
                               "2,3,4,7,8-pncdf" =  "lbxf03la.l2",
                               "1,2,3,4,7,8-hxcdf" =  "lbxf04la.l2",
                               "1,2,3,6,7,8-hxcdf" =  "lbxf05la.l2",
                               "1,2,3,4,6,7,8-hxcdf" =  "lbxf08la.l2",
                               "PCB 169" =  "lbxhxcla.l2",
                                "PCB 126" = "lbxpcbla.l2")) %>% 
  ggplot(aes(x = chem, y = Loading, fill = Group)) + geom_col() +
  facet_wrap(~ PC) + theme_bw() + 
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1),
        strip.background = element_rect(fill = "white")) +
  geom_hline(yintercept = 0, size = 0.2) +
  labs(x = "Chemicals",
       y = "Loadings")
```

## Principal Component Biplot w/ Log-Transformed PCA Results

```{r PCA biplot}
#Creates a biplot showing the loadings for principal component 1 and 2. 
autoplot(pca.ln, data = log.x, size = 0.8, colour = 'blue', alpha = 0.5,
         loadings = TRUE, loadings.colour = 'orange',
         loadings.label = TRUE, loadings.label.repel = T, 
         loadings.label.size = 2.5, loadings.label.colour = 'black',
         main = "Principal Component Analysis Biplot")
```

##Health Model

Combine PC scores with original data outcome and covariants

```{r}
x <- pca.ln$x #principal component scores for each chemical

pop <- read_csv(here::here("Data/studypop.csv")) %>% 
  clean_names() %>% 
  na.omit() %>%
  dplyr::select(-(3:10), -(17:26)) %>% 
  mutate(logtelomean = log(telomean))


pop_pca <- as.tibble((cbind(pop, x)))
```

Put 3 components (continuous) into a linear regression w/ covariates to estimate association with LTL.

```{r}
health_model_pca <- lm(logtelomean ~ PC1  + PC2 + PC3 +   
     lbxwbcsi + lbxlypct + lbxmopct + lbxnepct + lbxeopct + lbxbapct + age_cent + age_sq +
     as.factor(bmi_cat3) + as.factor(edu_cat) + as.factor(race_cat) + male, data = pop_pca)

summary(health_model_pca)
confint(health_model_pca) ##confidence intervals

health_model_pca %>% broom::tidy() %>% mutate(
  estimate = round(estimate, 4),
  std.error = round(std.error, 4),
  statistic = round(statistic, 4),
  p.value = round(p.value, 4))

health_model_pca %>% broom::glance() %>% mutate(p.value = round(p.value, 5))


PCA_betas_confint <- as.data.frame(c(health_model_pca$coefficients["PC1"], health_model_pca$coefficients["PC2"], health_model_pca$coefficients["PC3"])) %>% rownames_to_column() %>% 
  rename("variable" = 1) %>% 
  rename("beta" = 2) %>% 
  mutate('95%CI' = c("-0.018, -0.005", "-0.009, 0.011","-0.012, 0.017" ),
       'P-value' = c((summary(health_model_pca)$coefficients[2,4]),
                     (summary(health_model_pca)$coefficients[3,4]),
                     (summary(health_model_pca)$coefficients[4,4])))
PCA_betas_confint

#write.csv(PCA_betas_confint, file = "PCA_betas_confint.csv")

```

### Plot Model

```{r}
health_model_pca %>% broom::tidy() %>% as.tibble() %>% filter(grepl("PC", term)) %>% 
  ggplot(aes(x = term, y = estimate, color = term,
             ymin = estimate - 1.96*std.error,
             ymax = estimate + 1.96*std.error)) +
  geom_pointrange() + theme_bw() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  theme(legend.position = "none") + coord_flip() +
  labs(y = "Estimate", x = "Principal Component", title = "Association between Principal Component and LTL")
```
